Adaptative Computation:\\ An Essay in Memory of John H. Holland

### 2015-12-15 paper.tex
Submitted Draft
Submitted draft of paper.

% \begin{abstract}
% \input{abstract}
% \end{abstract}

@@

Holland is best known for his invention of \emph{genetic
  algorithms} (GAs), a family of search and optimization methods inspired
by biological evolution.  Since their invention in the 1960s, GAs have
inspired many related methods and led to the thriving subfield of
\emph{evolutionary computation}, with widespread scientific and
commercial applications.  While the mechanisms of GAs are well-known,
%Holland's broader motivations are less widely appreciated---to develop a 
%general theory of adaptation in complex systems.  
they were only one
offshoot of Holland's broader motivation---to develop a general
theory of adaptation in complex systems.  
%This motivation was the
%driving force in all of Holland's research.   
%to much of the computer science community, 
%the ideas that motivated Holland are less widely appreciated---to develop a general
%Holland's motivation 
%offshoot of Holland's much broader motivation--
%This motivation was the
%driving force in all of Holland's research.

In this short essay, we consider this larger framework, sketching the
recurring themes that were central to Holland's theory of
adaptive systems: (1) discovery and 
dynamics in adaptive search; (2) internal models and prediction; (3) 
exploratory modeling; and (4) universal properties of complex adaptive systems.  
%We discuss the role these
%themes have played in computer science, and highlight especially the
%ideas that we think remain relevant to today's research agendas.  [MM
%says: Are we going to do this?]

@@

Holland's goal of developing a general theory of adaptation 
%spurred
began with his early work on computer models of Hebbian learning
\cite{Rochester1956} and his reading of Ronald Fisher's classic book
that integrated genetics with Darwinian selection \cite{Fisher1930}.
As Holland read extensively in evolutionary biology,
economics, game theory, and control theory, he came to recognize that
adaptation was the most central concept in all these fields.  That is,
these fields all concern populations of agents that must continually
obtain information from uncertain changing environments and use it in
order to improve performance and enhance the chance of survival.

@@

\begin{itemize}
\item{\bf Populations, sampling, and implicit parallelism:}

% Adaptation occurs in
% populations of individuals evolving (or learning) over time, in
% which statistics can be leveraged to direct population dynamics.
%  The process of adaptive 
  Evolution can be framed as a search that leverages statistics to
  direct population dynamics.  A population can be thought of as a
  parallel sampling of many individuals (from the space of possible
  individuals).  Moreover, that same population can be thought of as
  implicitly sampling a much larger space of \emph{traits} that
  comprise those individuals.  Holland termed this implicit
  large-scale sampling of traits ``implicit parallelism.''
  Evolutionary dynamics biases these samples over time towards
  high-fitness regions of the search space.

%The population can be
%  thought of as a set of samples in a much larger search space, where
%  the processes of evolution
%  population can be thought of as a set of sa in which statistics 
%  can be leveraged to direct population dynamics.  (More on this below.)

\item{\bf Building blocks and recombination:} 
  In a population undergoing adaptation,
  individuals can be decomposed into \emph{building blocks}---sets of
  traits that are the evolutionary ``atoms'' of an individual's
  fitness or performance. (As an example from biology, Holland gives
  the \emph{Krebs cycle}, a core cellular metabolic
  pathway.)   Successful individuals
  are discovered in stages, first by finding useful
  building blocks through random sampling, and over
  time recombining such building blocks to create higher-fitness individuals.

@@

The question of how to balance exploitation and exploration---how to
optimally allocate trials to different arms based on their observed 
payoff---now becomes the question of how to optimally sample in the
vast space of possible building blocks, based on their estimated
contribution to fitness.  Of course, evolution deals in populations of
individuals, not building blocks.  There is no {\emph explicit} mechanism for
keeping statistics on how building blocks contribute to fitness. 
Holland's central idea here is that an approximation to
optimal building-block sampling does implicitly occur, as an emergent
property of evolutionary population dynamics.

@@

% SF liked 'icing on the cake.'

@@

Internal models are central to Holland's theory of adaptive systems.
He posits that all adaptive systems create and use internal models to
prosper in the face of ``perpetual novelty.'' 
%in continually changing environments.  
Models can be tacit and learned through evolutionary
time, as in the case of bacteria swimming up a chemical gradient, or
explicit and learned over a single lifespan, as in the case of
cognitive systems that incorporate experience into internal
representations through learning.  In Holland's view, the key activity
of an adaptive agent involves building and refining these data-driven
models of the environment.
%using them to make predictions and take actions,
%  and occasionally receiving positive or negative feedbacks
%  (intermittent rewards).

In his second book, \emph{Induction} \cite{Holland1989}, Holland and
his co-authors proposed inductive methods by which 
%tackled the question of how 
cognitive agents can
%learn 
construct---and profit from---internal models by combining
environmental inputs and rewards with stored knowledge.  In their
framework, a model defines a set of equivalence relations over
environmental states, together with a set of transition rules which
are learned over time based on environmental rewards (or punishments).
Models that form valid homomorphisms with the environment allow the
system to make accurate predictions.  In Holland's conception, the
equivalence classes are initially very general, for example, ``moving object''
and ``stationary object.'' Through experience and
learning, these can be specialized into more useful and precise subclasses, say
``insect'' and ``nest.''  Over time, the adaptive system builds up a
default hierarchy of rules covering general cases and refinements for
specific classes.

% For example, a general rule might
% state that, ``If X is a bird, then X can fly'', whereas a more specific
% one might state, ``If X is a bird with small wings and a large body,
% then X cannot fly'' \cite{Holland1989}.  When faced with a bird in
% the environment, both rules might compete to post their conclusion,
% but the more specific one would be favored if both of its conditions are met.

% Maybe this should be moved to section on Complexity
Although the idea of default hierarchies was prevalent in 
knowledge representation systems of the era, Holland made two key
contributions.  The first was his emphasis on homomorphisms as a formal way to
evaluate model validity, an idea that dates back to Ross Ashby's
\emph{An Introduction to Cybernetics} \cite{Ashby1956}. Holland's student
Bernard Ziegler developed this idea into a formal theory of computer modeling
and simulation \cite{Ziegler1976}.  
% %Today, even though computational modeling is a major
% %application area of computing, essential to most natural sciences and
% %engineering disciplines
Even today, these early homomorphic theories of modeling
stand as the most elegant approach we know of to characterize when
a model is consistent with the environment and how an intelligent
agent, human or artificial, can update a model to better reflect
reality.

Holland's second key contribution was describing a computational
mechanism, the \emph{learning classifier
  system}~\cite{Holland1986,Holland1977}, to illustrate how a cognitive system could
iteratively build up a detailed and hierarchical model of its
environment to enhance survival.  The key learning elements of this method, the
bucket-brigade algorithm combined with a genetic algorithm, presaged many of the ideas in modern reinforcement
learning.

@@

Holland's inspiration for classifier systems came from several
different disciplines, including Hebbian
learning, artificial intelligence, evolutionary biology, economics,
psychology, and control theory.
% MM says: We cite these elsewhere in the paper
%(e.g.,\cite{Bellman1961,Samuel1959}).  
Knowledge representation in the form of a
population of ``if-then'' rules seemed like a good choice, not only
because of its popularity in AI at the time but also from Holland's
early work on modeling Hebbian cell assemblies: ``In Hebb's view, a
cell assembly makes a simple statement: If such and such an event
occurs, then I will fire for a while at a high
rate'' \cite{Waldrop1993}. 
The if-then rules, when activated, compete to post their results on a
shared ``message list,'' serving as the system's short-term memory
(again inspired by Hebb's work and AI blackboard systems of the day).
Unlike blackboard systems, however, new rules are generated automatically
in a trial-and-error fashion.
% SF reverted this because lcs had several non-GA learning mechanisms,
% like 'covering'
%using a genetic algorithm.

Successful rules are strengthened over time if their predictions led
to positive rewards from the environment (and weakened otherwise)
through a credit-assignment method called the \emph{bucket-brigade}
algorithm, in which rules gaining rewards from the environment or from
other rules transfer some of their gains to those earlier-firing
``stage-setting'' rules that set up the conditions for the eventual
reward. Holland credited Arthur Samuel's pioneering work on machine
learning applied to checkers \cite{Samuel1959} as a key inspiration for these ideas.

%, e.g., finding a target, or avoiding a predator,

@@

%Finally, new rules for the population are discovered by a genetic
%algorithm, using the strength of rules as a fitness function.
%Classifier systems are fairly complex architectures, with learning
%occurring at multiple time scales.  
Although Holland proposed classifier systems as 
an executable theory of inductive processes in cognition, other researchers took it further, 
%other researchers took the
%computational system more seriously
applying it to areas as diverse
as poker-playing \cite{Smith1980}, control of gas pipeline
transmission \cite{Goldberg1983}, and modeling the stock market
\cite{Palmer1994}.  (See Ref.~\cite{Booker1989} for more details about
practical applications of classifier systems.)  Today, other
reinforcement learning methods are more popular for real-world
decision and control problems, but classifier systems can perhaps be
thought of as an essential stage-setting method that enabled the
development of later approaches.

@@

Holland's work focused on {\emph exploratory} models---those that explore basic principles and
mechanisms, even if they do not make specific or detailed
predictions.  Such models can show generically how certain behaviors
could be produced.  Holland pioneered a style of modeling that has
come to be known as ``individual-based'' or ``agent-based,'' in which
every component of a system is represented explicitly---e.g., every
trader in a stock market system or every cell in an immune system
model---and has a dynamic internal state. In such models, each agent has
its own behavior rules, which it can modify over time through learning.  In
order to capture the behavior of systems under spatial
constraints, these models are often defined over spatial structures,
such as networks or simple grids.

@@

%Similar to expert systems [MM says: ?]
%Such models are especially useful for studying systems
%that do not have analytic mathematical descriptions.  [MM took out to shorten this paragraph]
%Agent-based models can facilitate interdisciplinary collaborations because the
%underlying rules can be easily communicated.  
%[SF: THINKS this next text doesn't make sense until after complex
%systems seciton.]
The exploratory models
championed by Holland were idealized and not intended to provide detailed, domain-specific
predictions.  Instead they were meant to explore possible general
mechanisms of complex systems and thus provide insights that might
lead to more specific, detailed models.  Such idealized models are
akin to what Dennett has called ``intuition pumps''
\cite{Dennett1984}.

@@

It should be noted that Holland's view of modeling is by no means
typical.  For example, a textbook on computational modeling offers
 the following definition: ``Modeling is the application
of methods to analyze complex, real-world problems in order to make
predictions about what might happen with various actions''
\cite{Shiflet2014}.   This perspective rules out the
kind of exploratory modeling that Holland was most interested in.

@@

In the early 1990s, Holland teamed up with other Santa Fe Institute researchers,
including several economists, to tackle the mismatch between predictions
of rational expectations theory (the then dominant theory in economics) 
and empirically observed stock market behaviors. 
%that deviated from the predicted equilibrium solutions.  
In brief, most
economic theory of the day assumed that all participants in an economy
or financial market are 100\% rational and act to maximize their
individual gain.  In real life, however, actors in economies
and markets are rarely wholly rational, and financial markets often deviate
from rationality, for example with speculative bubbles and crashes.

@@

ECHO \cite{Holland1999,Hraber1997} was an even more ambitious exploratory model
that Holland and collaborators developed next. 
% ECHO's
% scope included the evolution of interacting populations of individual
% agents, which through competition and learning discover symbiotic
% triads such as the famous ant-fly-caterpillar interaction, the Wiksell
% triangle of trading relationships in economics, and the immune
% system's learned ability to distinguish `self' from `other.'
ECHO formalized Holland's idealization of complex adaptive systems into
a runnable computational system where agents evolved external markers
(called tags) and internal preferences, and then used them to acquire
resources, form higher
level aggregate structures (such as trading relationships, symbiotic groups,
trophic cascades, interdependent organizations, etc.).  ECHO agents
sometimes discovered mimicry and used it to deceive competitors.  
In many runs,  
the model developed increasingly complex structures and behaviors up
to a point.  When these runs stabilized, the resulting population of agents
% at least up to a point.   
was found to reproduce several well-known patterns 
observed in nature, most famously the rank-frequency distribution of
species diversity known as the Preston curve in ecology.
In layman's terms, ECHO evolved populations where 
%quantitative statement of the adage that
``most species are rare.''  
The broad scope of the model, together with its ability to produce
easily identifiable and well-known patterns from nature, was appealing to immunologists, economists, and evolutionary biologists alike.   Many
of the insights behind this project are described in Holland's third book \emph{Hidden Order} \cite{Holland1995}.

@@

Holland's later books, \emph{Emergence} \cite{Holland2000},
\emph{Signals and Boundaries} \cite{Holland2012}, and
\emph{Complexity: A Very Short Introduction} \cite{Holland2014} show how the
theories of adaptation that Holland developed during the earlier part
of his career fit into the larger landscape of complex systems
research.  Holland's focus on understanding the mechanisms by which complex patterns emerge and change,
rather than simply characterizing the patterns themselves (e.g.,
describing chaotic attractors or power laws), reflected his
determination to get to the heart of complex adaptive systems.
This determination represents the best of science.  Holland's
willingness to tackle the hardest questions, develop his own
formalisms, and use mathematics to provide insight sets a high bar 
scientists in all disciplines.
%to whch we
%all should aspire.

@@

John Holland was unusual in his ability to absorb the essence of other
disciplines, articulate grand overarching
principles, and them back them up with computational mechanisms and
mathematics.  Unlike most researchers he moved seamlessly among these
three modes of thinking, developing models that were years ahead of their time.  A
close reading of his work reveals the antecedents of many ideas prevalent in
machine learning today, e.g., reinforcement, learning in non-Markovian
environments and active learning.   His seminal genetic algorithm
spawned the field of evolutionary computation, and his insights
and wisdom helped define what are now referred to as the
\emph{sciences of complexity}.

Holland's many books and articles have influenced scientists around
the work and across many disciplines.  Closer to home, he introduced
several generations of students at the University of Michigan to
computation in natural systems, an idea that even today remains
somewhat controversial, despite successes in genetic algorithms for
engineering design, biomimicry for robotics, abstractions of phermone
trails in ant colonies for optimization methods, and the use of
mechanisms from immunology to improve computer security.

Behind the ideas, of course, is the man himself.   Everyone who knew
John personally will remember the gleam in his eye when he
encountered a new idea; his willingness to talk to anyone, no matter
how famous or how unfamous; and his incredible generosity and patience.
%His insights aboutwere deeper and more general than what often passes for work in biomimicry, e.g., for robots.
Holland's personality and humanity were somehow inextricably tangled
up with his intellectual contributions.  Since his death, many of
Holland's former students and colleagues have movingly described their
desire to emulate his personal qualities as much as his scientific
excellence.  John Holland's ideas, intellectual passion, and personal approach will serve as
beacons for research in intelligent and complex systems for many
years to come. 
%the world of science is poorer without John
%Holland on the scene.

>>>>

\begin{abstract}
In this essay we review the work and influence of the late John H. Holland, pioneer in the fields of evolutionary computation, machine learning, and complex systems.  While best known for his invention of \emph{genetic algorithms}, Holland's scientific motivation was much broader:  to develop a general theory of adaptation in complex systems.  We survey Holland's motivation and work with respect to four central themes, and describe how this work has had both broad and deep influence in modern approaches to computer science and complex systems science.  
\end{abstract}

@@

Holland is best known for his invention of \emph{genetic
  algorithms} (GAs), a family of search and optimization methods inspired
by biological evolution.  Since their invention in the 1960s, GAs have
inspired many related methods and led to the thriving subfield of
\emph{evolutionary computation}, with widespread scientific and
commercial applications.  While the mechanisms of GAs are well-known,
they were only one
offshoot of Holland's broader motivation---to develop a general
theory of adaptation in complex systems.

In this short essay, we consider this larger framework, sketching the
recurring themes that were central to Holland's theory of
adaptive systems: (1) discovery and 
dynamics in adaptive search; (2) internal models and prediction; (3) 
exploratory modeling; and (4) universal properties of complex adaptive systems.

@@

Holland's goal of developing a general theory of adaptation was
spurred both by his early work on computer models of Hebbian learning
\cite{Rochester1956} and his reading of Ronald Fisher's classic book
on integrating genetics with Darwinian selection \cite{Fisher1930}.
As Holland further read extensively in evolutionary biology,
economics, game theory, and control theory, he came to recognize that
adaptation was the most central concept in all these fields.  That is,
these fields all concern populations of agents that must continually
obtain information from uncertain, changing environments and use it in
order to improve performance and enhance the chance of survival.

@@

\begin{itemize}
\item{\bf Populations, sampling, and implicit parallelism:} 
  Evolution can be framed as a search that leverages statistics to
  direct population dynamics.  A population can be thought of as a
  parallel sampling of many individuals (from the space of possible
  individuals).  Moreover, that same population can be thought of as
  implicitly sampling a much larger space of \emph{traits} that
  comprise those individuals.  Holland termed this implicit
  large-scale sampling of traits ``implicit parallelism.''
  Evolutionary dynamics biases these samples over time towards
  high-fitness regions of the search space.

\item{\bf Building blocks and recombination:} 
  In a population undergoing adaptation,
  individuals can be decomposed into \emph{building blocks}---sets of
  traits that are the evolutionary ``atoms'' of an individual's
  fitness or performance. (As an example from biology, Holland gives
  the \emph{Krebs cycle}, a core cellular metabolic
  pathway.)   Successful individuals
  are discovered in stages, first by finding useful
  building blocks through stochastic sampling, and over
  time recombining such building blocks to create higher-fitness individuals.

@@

The question of balancing exploitation and exploration---how to
optimally allocate trials to different arms based on their observed 
payoffs---now becomes the question of how to optimally sample in the
vast space of possible building blocks, based on their estimated
contribution to fitness.  Of course, evolution deals in populations of
individuals, not building blocks.  There is no {\emph explicit} mechanism for
keeping statistics on how building blocks contribute to fitness. 
Holland's central idea here is that an approximation to
optimal building-block sampling does implicitly occur, as an emergent
property of evolutionary population dynamics.

@@

%The later, successful application of
%genetic algorithms to real-world optimization and learning tasks was,
%for Holland, just icing on the cake.
% SF liked 'icing on the cake.'

@@

Internal models are central to Holland's theory of adaptive systems.
He posits that all adaptive systems create and use internal models to
prosper in the face of ``perpetual novelty.''

Models can be tacit and learned through evolutionary time, as in the
case of bacteria swimming up a chemical gradient, or explicit and
learned over a single lifespan, as in the case of cognitive systems
that incorporate experience into internal representations through
learning.  In Holland's view, the key activity of an adaptive agent
involves building and refining these data-driven models of the
environment.

In his second book, \emph{Induction} \cite{Holland1989}, Holland and
his co-authors proposed inductive methods by which cognitive agents
can construct---and profit from---internal models by combining
environmental inputs and rewards with stored knowledge.  In their
framework, a model defines a set of equivalence relations---a
homomorphism---over environmental states, together with a set of
transition rules which are learned over time based on environmental
rewards (or punishments).  Models that form valid homomorphisms with
the environment allow the system to make accurate predictions.  In
Holland's conception, the equivalence classes are initially very
general, for example, ``moving object'' and ``stationary object.''
Through experience and learning, these can be specialized into more
useful and precise subclasses, say ``insect'' and ``nest.''  Over
time, the adaptive system builds up a default hierarchy of rules
covering general cases and refinements for specific classes.

Although the idea of default hierarchies was prevalent in knowledge
representation systems of the era, Holland made two key contributions.
The first was his emphasis on homomorphisms as a formal way to
evaluate model validity, an idea that dates back to Ross Ashby's
\emph{An Introduction to Cybernetics} \cite{Ashby1956}. Holland's
student Bernard Ziegler developed this idea into a formal theory of
computer modeling and simulation \cite{Ziegler1976}.  Even today,
these early homomorphic theories of modeling stand as the most elegant
approach we know of to characterize when a model is consistent with
the environment and how an intelligent agent, human or artificial, can
update a model to better reflect reality.

Holland's second key contribution was describing a computational
mechanism, the \emph{learning classifier
  system}~\cite{Holland1986,Holland1977}, to illustrate how a cognitive system could
iteratively build up a detailed and hierarchical model of its
environment to enhance survival.  The key learning elements of this method, the
\emph{bucket-brigade} algorithm combined with a genetic algorithm, presaged many of the ideas in modern reinforcement
learning.

@@

Holland's inspiration for classifier systems came from several
different disciplines, including Hebbian
learning, artificial intelligence, evolutionary biology, economics,
psychology, and control theory.
% MM says: We cite these elsewhere in the paper
%(e.g.,\cite{Bellman1961,Samuel1959}).  
Knowledge representation in the form of a population of ``if-then''
rules seemed like a good choice, not only because of its popularity in
AI at the time but also from Holland's early work on modeling Hebbian
cell assemblies: ``In Hebb's view, a cell assembly makes a simple
statement: If such and such an event occurs, then I will fire for a
while at a high rate'' \cite{Waldrop1993}.  The if-then rules, when
activated, compete to post their results on a shared ``message list,''
serving as the system's short-term memory (again inspired by Hebb's
work and AI blackboard systems of the day).  Unlike blackboard
systems, however, new rules are generated automatically in a
trial-and-error fashion, and can be selected and recombined via a
genetic algorithm.

Successful rules are strengthened over time if their predictions led
to positive rewards from the environment (and weakened otherwise)
through a credit-assignment method called the \emph{bucket-brigade}
algorithm, in which rules gaining rewards from the environment, or from
other rules, transfer some of their gains to earlier-firing
``stage-setting'' rules that set up the conditions for the eventual
reward. Holland credited Arthur Samuel's pioneering work on machine
learning applied to checkers \cite{Samuel1959} as a key inspiration for these ideas.

@@

%Finally, new rules for the population are discovered by a genetic
%algorithm, using the strength of rules as a fitness function.
%Classifier systems are fairly complex architectures, with learning
%occurring at multiple time scales.  
Although Holland proposed classifier systems as 
an executable theory of inductive processes in cognition, other researchers took it further, 
%other researchers took the
%computational system more seriously
applying it to areas as diverse as poker-playing \cite{Smith1980},
control of gas pipeline transmission \cite{Goldberg1983}, and modeling
the stock market \cite{Palmer1994}.  (See Ref.~\cite{Booker1989} for
more details about practical applications of classifier systems.)
Today, other reinforcement learning methods are more popular for
real-world decision and control problems, but classifier systems can
perhaps be thought of as an early stage-setting method that helped
build the foundations for later approaches.

@@

Holland's work focused on {\emph exploratory} models---those that explore basic principles and
mechanisms, even if they do not make specific or detailed
predictions.  Such models can show generically how certain behaviors
could be produced.  Holland pioneered a style of modeling that has
come to be known as ``individual-based'' or ``agent-based,'' in which
every component of a system is represented explicitly---e.g., every
trader in a stock market system or every cell in an immune system
model---and has a dynamic internal state. In such models, each agent has
its own behavior rules, which it can modify over time through learning.  In
order to capture the behavior of systems under spatial
constraints, these models are often defined over spatial structures
such as networks or simple grids.

@@

%Similar to expert systems [MM says: ?]
%Such models are especially useful for studying systems
%that do not have analytic mathematical descriptions.  [MM took out to shorten this paragraph]
%Agent-based models can facilitate interdisciplinary collaborations because the
%underlying rules can be easily communicated.  
%[SF: THINKS this next text doesn't make sense until after complex
%systems seciton.]
The exploratory models championed by Holland were not intended to
provide detailed, domain-specific predictions.  Instead they were
meant to explore possible general mechanisms of complex systems and
thus provide insights that might lead to more specific, detailed
models.  Such idealized models are akin to what philosopher Daniel Dennett has called
``intuition pumps'' \cite{Dennett1984}.

@@

%It should be noted that Holland's view of modeling is by no means
%typical.  For example, a textbook on computational modeling offers
% the following definition: ``Modeling is the application
%of methods to analyze complex, real-world problems in order to make
%predictions about what might happen with various actions''
%\cite{Shiflet2014}.   This perspective rules out the
%kind of exploratory modeling that Holland was most interested in.

@@

In the early 1990s, Holland teamed up with other Santa Fe Institute researchers,
including several economists, to tackle the mismatch between predictions
of rational expectations theory (the dominant theory in economics at the time) 
and empirically observed stock market behaviors. 
%that deviated from the predicted equilibrium solutions.  
In brief, most
economic theory of the day assumed that all participants in an economy
or financial market are fully rational, and act to maximize their
individual gain.  In real life, however, actors in economies
and markets are rarely wholly rational, and financial markets often deviate
from rationality, for example with speculative bubbles and crashes.

@@

ECHO \cite{Holland1999,Hraber1997} was an even more ambitious exploratory model
that Holland and collaborators developed during the 1990s. 
% ECHO's
% scope included the evolution of interacting populations of individual
% agents, which through competition and learning discover symbiotic
% triads such as the famous ant-fly-caterpillar interaction, the Wiksell
% triangle of trading relationships in economics, and the immune
% system's learned ability to distinguish `self' from `other.'
ECHO formalized Holland's idealization of complex adaptive systems
into a runnable computational system where agents evolved external
markers (called tags) and internal preferences, and then used them to
acquire resources, form higher-level aggregate structures, such as
trading relationships, symbiotic groups, trophic cascades, and
interdependent organizations.  ECHO agents sometimes discovered
mimicry and used it to deceive competitors.  In many runs, the model
developed increasingly complex structures and behaviors up to a point.
When these runs stabilized, the resulting population of agents was
found to reproduce several well-known patterns observed in nature,
most famously the rank-frequency distribution of species diversity
known as the Preston curve in ecology.  In layman's terms, ECHO
evolved populations where ``most species are rare.''

The broad scope of the model, together with its ability to produce
easily identifiable and well-known patterns from nature, was appealing
to immunologists, economists, and evolutionary biologists alike.  Many
of the insights behind this project are described in Holland's third
book, \emph{Hidden Order} \cite{Holland1995}.

@@

Holland's later books, \emph{Emergence} \cite{Holland2000},
\emph{Signals and Boundaries} \cite{Holland2012}, and
\emph{Complexity: A Very Short Introduction} \cite{Holland2014} show how the
theories of adaptation that Holland developed during the earlier part
of his career fit into the larger landscape of complex systems
research.  Holland's focus on understanding the mechanisms by which complex patterns emerge and change,
rather than simply characterizing the patterns themselves (e.g.,
describing chaotic attractors or power laws), reflected his
determination to get to the heart of complex adaptive systems.
This determination represents the best of science.  Holland's
willingness to tackle the hardest questions, develop his own
formalisms, and use mathematics to provide insight sets a high bar for
scientists in all disciplines.
%to whch we
%all should aspire.

@@

John Holland was unusual in his ability to absorb the essence of other
disciplines, articulate grand overarching
principles, and them back them up with computational mechanisms and
mathematics.  Unlike most researchers, Holland moved seamlessly among these
three modes of thinking, developing models that were years ahead of their time.  A
close reading of his work reveals the antecedents of many ideas prevalent in
machine learning today, e.g., reinforcement, learning in non-Markovian
environments and active learning.   His seminal genetic algorithm
spawned the field of evolutionary computation, and his insights
and wisdom helped define what are now referred to as the
\emph{sciences of complexity}.

Holland's many books and articles have influenced scientists around
the world and across many disciplines.  Closer to home, he introduced
several generations of students at the University of Michigan to
computation in natural systems, an idea that even today remains
somewhat controversial, despite successes in genetic algorithms for
engineering design, biomimicry for robotics, abstractions of pheromone
trails in ant colonies for optimization methods, and the use of
mechanisms from immunology to improve computer security.

Behind the ideas, of course, is the man himself.   Everyone who knew
John personally will remember the gleam in his eye when he
encountered a new idea; his willingness to talk to anyone, no matter
how famous or obscure; and his incredible generosity and patience.
%His insights aboutwere deeper and more general than what often passes for work in biomimicry, e.g., for robots.
Holland's personality and humanity were somehow inextricably tangled
up with his intellectual contributions.  Since his death, many of
Holland's former students and colleagues have movingly described their
desire to emulate his personal qualities as much as his scientific
excellence.  John Holland's ideas, intellectual passion, and personal approach will serve as
beacons for research in intelligent and complex systems for many
years to come. 
%the world of science is poorer without John
%Holland on the scene.
